Bioconductor workflow for single-cell RNA sequencing: Normalization, dimensionality reduction, clustering, and lineage inference

### 2017-07-04 workflow/workflow.Rmd
Edits to text.

Single-cell RNA sequencing (scRNA-seq) is a powerful and promising class of high-throughput assays that enable researchers to measure genome-wide transcription levels at the resolution of single cells. To properly account for features specific to scRNA-seq, such as zero inflation and high levels of technical noise, several novel statistical methods have been developed to tackle questions that include normalization, dimensionality reduction, clustering, the inference of cell lineages and pseudotimes, and differential expression (DE) analysis. While each individual method is useful on its own for addressing a specific question, there is an increasing need for workflows that integrate these tools to yield a seamless scRNA-seq data analysis pipeline. This is all the more true, with novel sequencing technologies that allow an increasing number of cells to be sequenced in each run. For example, the Chromium Single Cell 3' Solution was recently used to sequence and profile about 1.3 million cells from embryonic mouse brains.

@@

Here, we propose an integrated workflow for dowstream analysis, with the following four main steps: (1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates, using the `zinbwave` Bioconductor package; (2) robust and stable cell clustering using resampling-based sequential ensemble clustering implemented in the `clusterExperiment` Bioconductor package; (3) inference of cell lineages and ordering of the cells by developmental progression along lineages, using the `slingshot` R package; and (4) DE analysis along lineages. Throughout the workflow, we use a unique `SummarizedExperiment` object to store the scRNA-seq data along with any metadata available from the experiment.

@@

```{r packages}
# Bioconductor:
library(clusterExperiment)
library(BiocParallel)
library(zinbwave)
library(scone)

@@

# CRAN
library(doParallel)
library(RColorBrewer)
library(gam)
set.seed(20)
```

Note that in order to successfully run the workflow, we need the devel versions
of the Bioconductor packages `scone (>=1.1.2)`, `zinbwave (>=0.99.5)` and `clusterExperiment (>=1.3.2)` [DR: Elizabeth, please check!]. We recommend running Bioconductor 3.6 (currently the devel version). See https://www.bioconductor.org/developers/how-to/useDevel/ on how to use the Bioconductor devel.

@@

We remove the ERCC spike-in sequences and the CreER gene, as the latter corresponds to the estrogen receptor fused to Cre recombinase (Cre-ER), which is used to activate HBC into differentiation following injection of tamoxifen (see [@Fletcher2017] for details).

@@

Using the Bioconductor R package `scone`, we remove low-quality cells that do not pass our quality control filter (See Figure \@ref(fig:scone)). The function `metric_sample_filter` uses several criteria to filter out low-quality samples. First, we filter out samples with low total number of reads or low alignment percentages. Then, we look at the number of housekeeping genes that are detected in each sample, and filter out those samples with a low detection rate. See the [scone vignette](https://www.bioconductor.org/packages/release/bioc/vignettes/scone/inst/doc/sconeTutorial.html) for details on the filtering procedure.

@@

```{r}
# Filtering to top 1000 most variable genes
vars <- rowVars(log1p(assay(se)))
names(vars) <- rownames(se)
vars <- sort(vars, decreasing = TRUE)
core <- se[names(vars)[1:1000],]
```

@@

Overall, after the pre-processing steps, our dataset has 1,000 genes and 747 cells.

@@

The function `zinbwave` returns a `SummarizedExperiment` object that includes normalized expression measures, defined as deviance residuals from the fit of the ZINB-WaVE model, with user-supplied gene- and cell-level covariates. Such residuals can be used for visualization purposes (e.g., in heatmaps, boxplots). Note that in this case the low-dimensional matrix `W` is not included in the residuals computation, to avoid the removal of the biological signal of interest.

@@

The next step of the workflow is to cluster the cells according to the low-dimensional matrix `W` computed in the previous step. We use the resampling-based sequential ensemble clustering (RSEC) frameworkd implemented in the `RSEC` function from the Bioconductor R package `clusterExperiment`. Specifically, given a set of user-supplied base clustering algorithms and associated tuning parameters (e.g., _k_-means, with a range of values for _k_), RSEC generates a collection of candidate clusterings, with the option of resampling cells and using a sequential tight clustering procedure as in [@Tseng2005]. A consensus clustering is obtained based on the levels of co-clustering of samples across the candidate clusterings. The consensus clustering is further condensed by merging similar clusters, which is done by creating a hierarchy of clusters, working up the tree, and testing for differential expression between sister nodes, with nodes of insufficient DE collapsed. As in supervised learning, resampling greatly improves the stability of clusters and considering an ensemble of methods and tuning parameters allows us to capitalize on the different strengths of the base algorithms and avoid the subjective selection of tuning parameters.

@@

We now demonstrate how to use the R software package `slingshot` to infer branching cell lineages and order cells by developmental progression along each lineage. The method, proposed in [@Street2017], comprises two main steps: (1) The inference of the global lineage structure (i.e., the number of lineages and where they branch) using a minimum spanning tree (MST) on the clusters identified above by `RSEC` and (2) the inference of cell pseudotime variables along each lineage using a novel method of simultaneous principal curves. The approach in (1) allows us to identify any number of novel lineages while also accommodating the use of domain-specific knowledge to supervise parts of the tree (e.g., known terminal states); the approach in (2) yields robust pseudotimes for smooth, branching lineages.

@@

To infer lineages and pseudotimes, we use the package `slingshot`, with input the 4-dimensional MDS of the low-dimensional matrix `W`. We found that the slingshot results were robust to the number of dimensions _k_ for the MDS (we tried _k_ from 2 to 5). Here, we use the unsupervised version of `slingshot`, where we only provide the identity of the start cluster but not of the end clusters.

@@

Before fitting the simultaneous principal curves, we check the structure of the lineages by plotting the MST on the clusters. This shows that our implementation has recovered the lineages found in the published work (Figure \@ref(fig:tree)). The `slingshot` package also includes methods for 3-dimensional visualizations similar to \@ref(fig:stemcelldiff), using the `plot3d` function from the package `rgl`.

@@

Having found the global lineage structure, we now construct a set of smooth, branching curves in order to infer the pseudotime variables. Simultaneous principal curves are constructed from the individual cells along each lineage, rather than the cell clusters. This makes them more stable and better suited for assigning cells to lineages. The final curves are shown in \@ref(fig:curves).

>>>>

Single-cell RNA sequencing (scRNA-seq) is a powerful and promising class of high-throughput assays that enable researchers to measure genome-wide transcription levels at the resolution of single cells. To properly account for features specific to scRNA-seq, such as zero inflation and high levels of technical noise, several novel statistical methods have been developed to tackle questions that include normalization, dimensionality reduction, clustering, the inference of cell lineages and pseudotimes, and the identification of differentially expressed (DE) genes. While each individual method is useful on its own for addressing a specific question, there is an increasing need for workflows that integrate these tools to yield a seamless scRNA-seq data analysis pipeline. This is all the more true, with novel sequencing technologies that allow an increasing number of cells to be sequenced in each run. For example, the Chromium Single Cell 3' Solution was recently used to sequence and profile about 1.3 million cells from embryonic mouse brains.

@@

Here, we propose an integrated workflow for dowstream analysis, with the following four main steps: (1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates, using the `zinbwave` Bioconductor package; (2) robust and stable cell clustering using resampling-based sequential ensemble clustering, as implemented in the `clusterExperiment` Bioconductor package; (3) inference of cell lineages and ordering of the cells by developmental progression along lineages, using the `slingshot` R package; and (4) DE analysis along lineages. Throughout the workflow, we use a unique `SummarizedExperiment` object to store the scRNA-seq data along with any metadata available from the experiment.

@@

```{r packages}
# Bioconductor
library(BiocParallel)
library(clusterExperiment)
library(scone)
library(zinbwave)

@@

# CRAN
library(doParallel)
library(gam)
library(RColorBrewer)

set.seed(20)
```

Note that in order to successfully run the workflow, we need the devel versions
of the Bioconductor packages `scone (>=1.1.2)`, `zinbwave (>=0.99.5)`, and `clusterExperiment (>=1.3.2)` [DR: Elizabeth, please check!]. We recommend running Bioconductor 3.6 (currently the devel version; see https://www.bioconductor.org/developers/how-to/useDevel/).

@@

We remove the ERCC spike-in sequences and the CreER gene, as the latter corresponds to the estrogen receptor fused to Cre recombinase (Cre-ER), which is used to activate HBCs into differentiation following injection of tamoxifen (see [@Fletcher2017] for details).

@@

Using the Bioconductor R package `scone`, we remove low-quality cells according to the quality control filter implemented in the function `metric_sample_filter` and based on the following criteria (Figure \@ref(fig:scone)). First, we filter out samples with low total number of reads or low alignment percentage. Next, we filter out samples with a low detection rate for housekeeping genes. See the [scone vignette](https://www.bioconductor.org/packages/release/bioc/vignettes/scone/inst/doc/sconeTutorial.html) for details on the filtering procedure.

@@

```{r}
# Filtering to top 1,000 most variable genes
vars <- rowVars(log1p(assay(se)))
names(vars) <- rownames(se)
vars <- sort(vars, decreasing = TRUE)
core <- se[names(vars)[1:1000],]
```

@@

Overall, after the above pre-processing steps, our dataset has 1,000 genes and 747 cells.

@@

The function `zinbwave` returns a `SummarizedExperiment` object that includes normalized expression measures, defined as deviance residuals from the fit of the ZINB-WaVE model with user-supplied gene- and cell-level covariates. Such residuals can be used for visualization purposes (e.g., in heatmaps, boxplots). Note that, in this case, the low-dimensional matrix `W` is not included in the computation of residuals to avoid the removal of the biological signal of interest.

@@

The next step of the workflow is to cluster the cells according to the low-dimensional matrix `W` computed in the previous step. We use the resampling-based sequential ensemble clustering (RSEC) framework implemented in the `RSEC` function from the Bioconductor R package `clusterExperiment`. Specifically, given a set of user-supplied base clustering algorithms and associated tuning parameters (e.g., _k_-means, with a range of values for _k_), RSEC generates a collection of candidate clusterings, with the option of resampling cells and using a sequential tight clustering procedure as in [@Tseng2005]. A consensus clustering is obtained based on the levels of co-clustering of samples across the candidate clusterings. The consensus clustering is further condensed by merging similar clusters, which is done by creating a hierarchy of clusters, working up the tree, and testing for differential expression between sister nodes, with nodes of insufficient DE collapsed. As in supervised learning, resampling greatly improves the stability of clusters and considering an ensemble of methods and tuning parameters allows us to capitalize on the different strengths of the base algorithms and avoid the subjective selection of tuning parameters.

@@

We now demonstrate how to use the R software package `slingshot` to infer branching cell lineages and order cells by developmental progression along each lineage. The method, proposed in [@Street2017], comprises two main steps: (1) The inference of the global lineage structure (i.e., the number of lineages and where they branch) using a minimum spanning tree (MST) on the clusters identified above by `RSEC` and (2) the inference of cell pseudotime variables along each lineage using a novel method of simultaneous principal curves. The approach in (1) allows the identification of any number of novel lineages while also accommodating the use of domain-specific knowledge to supervise parts of the tree (e.g., known terminal states); the approach in (2) yields robust pseudotimes for smooth, branching lineages.

@@

To infer lineages and pseudotimes, we apply Slingshot to the 4-dimensional MDS of the low-dimensional matrix `W`. We found that the Slingshot results were robust to the number of dimensions _k_ for the MDS (we tried _k_ from 2 to 5). Here, we use the unsupervised version of Slingshot, where we only provide the identity of the start cluster but not of the end clusters.

@@

Before fitting the simultaneous principal curves, we check the structure of the lineages by plotting the MST on the clusters. This shows that our implementation has recovered the lineages found in the published work (Figure \@ref(fig:tree)). The `slingshot` package also includes methods for 3-dimensional visualizations as in Figure \@ref(fig:stemcelldiff), using the `plot3d` function from the package `rgl`.

@@

Having found the global lineage structure, we now construct a set of smooth, branching curves in order to infer the pseudotime variables. Simultaneous principal curves are constructed from the individual cells along each lineage, rather than the cell clusters. This makes them more stable and better suited for assigning cells to lineages. The final curves are shown in Figure \@ref(fig:curves).
